---
title: "Statistical Appendix: Methods, Formulas & Worked Examples"
author: "Vid Kuder Marušič"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Statistical Appendix: Methods, Formulas & Worked Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  warning = FALSE
)
library(vidternary)
```

# Purpose of this document

This appendix documents every mathematical/statistical method used in the
**vidternary** application: what question it answers, the exact formula
implemented, the assumptions it relies on, how to read the result, and —
critically — **when the result cannot be trusted**. Every worked example
below runs the *actual* package functions on a small, hand-built ("toy")
dataset, so the numbers you see are not hand-typed illustrations - they are
live output, guaranteed to match what the app itself computes. If a formula
in the app ever changes, this document's numbers change with it the next
time it is rebuilt.

For a practical, click-by-click guide to the application's tabs, see the
companion **User Guide** vignette. This document assumes you already know
*where* each method lives in the interface and focuses purely on the
*statistics*.

Companion glossary terms used throughout: **NMI** = non-metallic inclusion,
**ECD** = equivalent circle diameter (the diameter of a circle with the
same area as the inclusion), **Wt%** = weight percent (chemical
composition), **control area** = one SEM field-of-view (or an artificial
equal-sized slice of the data, see §6).

---

# Ternary coordinates (barycentric normalization)

**What it answers:** how does a data point's composition split across
three chosen quantities (e.g. three oxide groups), positioned inside a
triangle where each corner represents 100% of one quantity?

**Formula:** for values $A$, $B$, $C$ (already summed across whichever
columns were assigned to each corner),

$$a = \frac{A}{A+B+C}, \quad b = \frac{B}{A+B+C}, \quad c = \frac{C}{A+B+C}$$

so that $a+b+c=1$ always, regardless of the original units. This is a
purely geometric normalization - there is no "significance" to test, and
no failure mode other than $A+B+C=0$ (undefined; the app drops such rows).

**Worked example**

```{r ternary-example}
# A fictional inclusion measured at Al2O3 = 20, SiO2 = 50, CaO = 30 (wt%,
# not required to sum to 100 - the app re-normalizes automatically)
A <- 20; B <- 50; C <- 30
total <- A + B + C
coords <- c(a = A / total, b = B / total, c = C / total)
coords
stopifnot(abs(sum(coords) - 1) < 1e-12)
```

```{r ternary-plot, fig.height=5}
if (requireNamespace("Ternary", quietly = TRUE)) {
  Ternary::TernaryPlot(atip = "Al2O3", btip = "SiO2", ctip = "CaO",
                        grid.lines = 5, grid.lty = "dotted")
  Ternary::TernaryPoints(list(coords), col = "#357ABD", pch = 16, cex = 1.5)
}
```

**Interpreting the result:** the point's position shows relative
proportions only - a point near the SiO2 corner means SiO2 dominates
*among these three quantities*, not that SiO2 is high in absolute terms.

**When it can't help you:** if two of the three corner values are
routinely near zero, all points crowd into one corner and the diagram
becomes uninformative - reconsider which quantities you assign to A/B/C.

---

# Descriptive statistics & coefficient of variation

**What it answers:** the basic shape of one numeric column's distribution.

**Formulas:** mean $\bar{x}$, median, standard deviation $s$, and the
coefficient of variation $CV\% = 100 \times s / \bar{x}$ - CV is the one
non-obvious addition here: it expresses spread *relative to* the mean, so
variables on very different scales (e.g. a trace element at 0.01 wt% vs.
Fe at 90 wt%) become comparable.

**Worked example** (five fictional ECD measurements, µm)

```{r desc-stats-example}
ecd <- c(0.8, 1.2, 0.9, 2.5, 1.1)
df_toy <- data.frame(ECD = ecd)
build_descriptive_stats_table(df_toy)
```

**Interpreting the result:** a CV% above roughly 50-100% signals a
right-skewed distribution typical of inclusion-size data (most inclusions
small, a few very large) - this is expected, not a data-quality problem,
and is exactly why EVS (§9) exists as a dedicated method for the tail
rather than relying on the mean.

**When it can't help you:** with fewer than ~5-10 observations the mean/SD
are unstable - treat descriptive statistics on very small groups as
indicative only, not conclusive.

---

# Correlation analysis (Pearson's r)

**What it answers:** whether two numeric variables move together linearly.

**Formula:**

$$r = \frac{\sum (x_i-\bar{x})(y_i-\bar{y})}{\sqrt{\sum(x_i-\bar{x})^2}\sqrt{\sum(y_i-\bar{y})^2}}, \qquad -1 \le r \le 1$$

**Worked example** (fictional Al vs. Si wt% across six samples, designed to
show a strong negative relationship)

```{r correlation-example}
df_corr <- data.frame(Al = c(1.2, 1.8, 2.1, 0.9, 2.6, 1.5),
                       Si = c(8.5, 6.9, 6.0, 9.2, 5.1, 7.8))
build_correlation_pairs_table(df_corr)
```

```{r correlation-plot, fig.height=4}
invisible(create_correlation_plot(df_corr, title = "Toy example"))
```

**Interpreting the result:** $|r| > 0.7$ is conventionally "strong",
$0.3-0.7$ "moderate", below $0.3$ "weak" - but always look at the scatter,
not just $r$: Pearson's r only captures *linear* relationships and is
distorted by outliers.

**When it can't help you:** compositional (Wt%) columns are constrained to
sum to a constant, which mechanically induces spurious negative
correlations between them (the "closure problem") - see §12 (CoDA) for the
correct fix before trusting correlations between raw composition columns.

---

# IQR, Z-score and MAD outlier filters

These three appear together in the app's "Statistical Filtering" section
because they answer the same question - *"is this value unusually high?"*
- with different assumptions about the data's shape. **All three flag
positive (high-value) outliers only** in this app, since the target use
case is spotting unusually large inclusions, not unusually small ones.

## IQR method

**Formula:** flag $x$ if $x > Q_3 + 1.5 \times IQR$, where $IQR = Q_3 - Q_1$.
Distribution-free (no normality assumption); the classic exploratory-data
-analysis convention (Tukey, 1977).

## Z-score method

**Formula:** flag $x$ if $z = (x-\bar{x})/s > 3$. Assumes the bulk of the
data is roughly normal - the mean and SD it's built on are themselves
pulled around by the very outliers being searched for.

## MAD method

**Formula:** flag $x$ if $x > \text{median} + 3 \times 1.4826 \times MAD$,
where $MAD = \text{median}(|x_i - \text{median}(x)|)$ and $1.4826$ rescales
MAD to be comparable to SD under normality. Unlike the Z-score, both the
median and MAD are themselves robust to outliers, so this method doesn't
suffer the same self-distortion.

**Worked example** (six fictional ECD values with one obvious outlier)

```{r outlier-filters-example}
ecd2 <- c(0.50, 0.60, 0.55, 0.62, 0.58, 5.20)
df_out <- data.frame(ECD = ecd2)

cat("IQR flags:   ", get_iqr_outlier_flags(df_out, "ECD"), "\n")
cat("Z-score flags:", get_zscore_outlier_flags(df_out, "ECD"), "\n")
cat("MAD flags:    ", get_mad_outlier_flags(df_out, "ECD"), "\n")

q3 <- quantile(ecd2, 0.75); iqr <- IQR(ecd2)
cat(sprintf("IQR upper bound: %.3f\n", q3 + 1.5 * iqr))
cat(sprintf("Z-score of the outlier: %.2f\n", (5.20 - mean(ecd2)) / sd(ecd2)))
cat(sprintf("MAD upper bound: %.3f\n", median(ecd2) + 3 * mad(ecd2)))
```

**Interpreting the result:** all three agree on this obvious case. They
diverge most on *moderate* outliers in small samples - if the three
methods disagree, that itself is informative (the point is borderline, not
unambiguously anomalous).

**When they can't help you:** with $n<10$ or so, $Q_1$/$Q_3$/SD are too
unstable to trust literally; and none of the three know anything about
*multiple* variables jointly - a point can be unremarkable on every single
column and still be a joint outlier (see Mahalanobis / Isolation Forest,
next).

---

# Mahalanobis distance

**What it answers:** how far is a point from a reference distribution's
center, accounting for the correlation structure between variables (unlike
Euclidean distance, which treats every variable as independent and
equally scaled)?

**Formula:**

$$MD(x) = \sqrt{(x-\mu)^\top \Sigma^{-1} (x-\mu)}$$

where $\mu$ and $\Sigma$ are the mean vector and covariance matrix of the
*reference* dataset. Squared Mahalanobis distance of multivariate-normal
data asymptotically follows a $\chi^2$ distribution with $p$ degrees of
freedom ($p$ = number of variables) - this gives the classic 95%/99%
thresholds reported alongside the app's own threshold.

**App-specific threshold:** rather than the classic $\chi^2$ cutoff, the
default automatic threshold uses

$$MD_{thresh} = MD_{mean} + \sqrt{\frac{100}{100+\lambda-\omega}} \times MD_{std}$$

where $\lambda$ (strictness) and $\omega$ (leniency) are tunable
sensitivity parameters from Vode et al., 2022
(<https://doi.org/10.3390/ma15020684>).

**Worked example** (two toy chemistry columns; reference cluster around
Al≈2, Si≈5, plus one target point far outside it)

```{r mahalanobis-example}
set.seed(1)
reference <- data.frame(Al = rnorm(30, 2, 0.4), Si = rnorm(30, 5, 0.6))
target <- data.frame(Al = c(2.1, 1.9, 2.3, 8.5), Si = c(5.2, 4.8, 5.5, 1.0))
# row 4 is the planted outlier

res_md <- compute_mahalanobis_distance(
  target, reference, lambda = 1, omega = 0,
  selected_columns = c("Al", "Si"), mdthresh_mode = "auto"
)
data.frame(row = 1:4, MD = round(res_md$distances, 2),
           flagged_as_outlier = res_md$outlier_indices)
cat(sprintf("\nThreshold (auto, λ=1, ω=0): %.2f\n", res_md$MDthresh))
cat(sprintf("95%% chi-square reference threshold (df=%d): %.2f\n",
            res_md$df, qchisq(0.95, df = res_md$df)))
```

**Interpreting the result:** the planted outlier's MD is far above both
the app's threshold and the classic $\chi^2_{0.95}$ reference - the other
three rows, drawn from the same distribution as the reference, correctly
stay below threshold.

**When it can't help you:**

- Needs at least $2\times$ as many observations as variables in *both*
  target and reference (the app enforces this).
- Assumes the reference data is roughly elliptically distributed
  (multivariate-normal-*ish*); a reference set that's itself bimodal or
  contains its own outliers will distort $\mu$ and $\Sigma$, making the
  distance measure unreliable for everyone.
- Highly correlated (>0.9) selected columns make the covariance matrix
  near-singular - the app warns about this during validation.

---

# Isolation Forest

**What it answers:** the same "is this point anomalous?" question as
Mahalanobis, but without assuming any particular distribution shape or
linear correlation structure - it works by how *easily* a point can be
isolated by random splits (Liu, Ting & Zhou, 2008,
<https://doi.org/10.1109/ICDM.2008.17>).

**How it works (conceptually):** many random binary-split trees are built
on the reference data. Points that sit in sparse regions of the data get
isolated (separated into their own leaf) after very few splits; points
embedded in dense regions take many splits. Average path length across all
trees becomes the anomaly score - short average path = anomalous.

**Worked example** (same toy chemistry data as the Mahalanobis example)

```{r isolation-forest-example}
res_if <- compute_isolation_forest(target, reference, selected_columns = c("Al", "Si"),
                                    contamination = 0.10, seed = 1)
data.frame(row = 1:4, score = round(res_if$scores, 3),
           flagged_as_outlier = res_if$outlier_indices)
cat(sprintf("\nThreshold (top %.0f%% of reference scores): %.3f\n",
            res_if$contamination * 100, res_if$threshold))
```

**Interpreting the result:** the same planted outlier scores clearly above
threshold, matching Mahalanobis here - agreement between the two methods
is reassuring; when they *disagree*, it usually means the anomaly is
non-linear/non-elliptical (something Mahalanobis can miss but Isolation
Forest can catch).

**When it can't help you:** the `contamination` parameter is essentially a
prior belief about what fraction of the reference is "normal" - set it too
high and normal points get flagged; too low and real anomalies slip
through. With very small reference samples (well under 100 rows) tree
splits become unstable and scores noisy.

---

# Extreme Value Statistics (Murakami's √area method)

**What it answers:** given the sizes of the *largest* inclusion in each of
several inspected regions, what size of inclusion should be expected if a
much larger area were inspected? Standard method for steel cleanliness
rating, standardized in ASTM E2283 and originating with
Murakami, 1994 (<https://doi.org/10.6028/jres.099.032>).

**Method:**

1. Split the inspected area into $n$ equal **control areas** $S_0$.
2. Take the largest inclusion (by $\sqrt{\text{Area}}$) in each - the
   **block maximum**.
3. Assign each of the $n$ sorted block maxima a plotting position
   $F_j = j/(n+1)$ and a reduced Gumbel variate $y_j = -\ln(-\ln F_j)$.
4. Fit a straight line $\sqrt{Area} = a + b\cdot y$ by least squares.
5. To predict the expected largest inclusion over $T$ control areas (the
   **return period**), solve for $y_T = -\ln(-\ln(1-1/T))$ and read off
   the fitted line, with a 95% prediction interval.

**Worked example** (six fictional control areas, each with a few
inclusions; sizes in µm)

```{r evs-example}
set.seed(2)
toy_incl <- data.frame(
  field = rep(1:6, times = c(4, 3, 5, 4, 3, 4)),
  sqrt_area = c(runif(4, 2, 8), runif(3, 2, 9), runif(5, 2, 7),
                runif(4, 2, 10), runif(3, 2, 6), runif(4, 2, 9))
)
block_max <- compute_block_maxima(toy_incl, area_col = "sqrt_area", group_col = "field")
# NB: compute_block_maxima expects an AREA column (it takes sqrt() itself);
# here sqrt_area is already sqrt(area)-like for a compact toy example, so
# square it first to demonstrate the real input contract:
toy_incl$area <- toy_incl$sqrt_area^2
block_max <- compute_block_maxima(toy_incl, area_col = "area", group_col = "field")
block_max

fit <- fit_evs_gumbel(block_max$sqrt_area_max)
cat(sprintf("Fitted line: sqrt(Area) = %.3f + %.3f * y   (R^2 = %.3f)\n",
            fit$intercept, fit$slope, fit$r_squared))

pred <- predict_evs_max(fit, return_period = 50)
cat(sprintf("Predicted largest inclusion over T=50 control areas: %.2f um [%.2f, %.2f] 95%% PI\n",
            pred$predicted, pred$lower, pred$upper))
```

```{r evs-plot}
print(create_gumbel_plot(fit, pred))
```

**Interpreting the result:** $R^2$ close to 1 means the block maxima line
up well on the Gumbel probability plot - a good sign the extreme-value
model fits. The prediction interval widens quickly as $T$ grows (predicting
far beyond the inspected area is inherently uncertain) - report the
interval, not just the point prediction.

**When it can't help you:**

- Needs **at least 3 control area groups** to fit anything at all, and
  realistically wants many more (10+) for a stable line.
- $R^2$ alone doesn't prove Gumbel-appropriateness - see §10
  (Anderson-Darling) for a real goodness-of-fit test.
- **$T$'s physical meaning depends entirely on how control areas were
  defined**: if grouped by a real field/frame-of-view ID, $T=100$ means a
  real area 100× one field. If groups were an artificial "split into N
  equal chunks" fallback (no real field ID available), $T=100$ only means
  "100× as many statistical groups" - not a calibrated physical area. The
  app surfaces this distinction explicitly in the EVS tab.

---

# Anderson-Darling goodness-of-fit test (for the Gumbel fit)

**What it answers:** do the block maxima *actually* follow a Gumbel
distribution, or does the EVS line above merely look straight-ish while
hiding real deviation (e.g. two inclusion populations with different size
distributions mixed together)?

**Formula:** the Anderson-Darling statistic

$$A^2 = -n - \frac{1}{n}\sum_{i=1}^{n}(2i-1)\big[\ln F(x_i) + \ln(1-F(x_{n+1-i}))\big]$$

weights deviations in the *tails* of the distribution more heavily than
the middle - exactly where an inclusion-rating extrapolation is most
sensitive to mismatch.

**Why this app doesn't use tabulated critical values:** published A-D
critical values for the Gumbel distribution
(Anderson & Darling, 1952,
<https://doi.org/10.1214/aoms/1177729437>; Stephens, 1977,
<https://doi.org/10.1093/biomet/64.3.583>) assume the Gumbel parameters
were fit by maximum likelihood. This app instead fits by least-squares
regression on the probability plot (§9) - a different estimator with a
different null distribution for $A^2$. Using the MLE-calibrated table
directly was checked empirically (1000 simulated true-Gumbel samples) and
over-rejected a true Gumbel fit at ~15% instead of the intended 5%.
Instead, the app **simulates the correct null distribution directly** by a
parametric bootstrap: repeatedly generate synthetic samples from the
*fitted* Gumbel model, refit each the same least-squares way, and see
where the observed $A^2$ falls among them.

**Worked example** (reusing the EVS fit above)

```{r ad-test-example}
gof <- gumbel_goodness_of_fit(fit, n_sim = 999, seed = 42)
cat(sprintf("A^2 = %.3f, bootstrap p-value %s\n", gof$statistic, gof$p_value_bracket))
cat(sprintf("Reject Gumbel fit at 5%% level? %s\n", gof$reject_at_05))
```

**Interpreting the result:** $p \ge 0.05$ (the toy example's, since the
data really was generated close to uniform block maxima) means no evidence
against Gumbel - proceed with the EVS prediction. $p<0.05$ means the block
maxima measurably deviate from a straight Gumbel line - a real warning
sign, not a false alarm.

**When it can't help you:** with very few control areas (under ~8-10) the
bootstrap has little power to detect real deviation - a non-significant
result there is weak reassurance, not strong evidence of a good fit.

---

# Clark-Evans nearest-neighbour test (spatial clustering)

**What it answers:** are inclusions randomly scattered across the
inspected area, clustered together (e.g. banding, a localized defect
zone), or more evenly spread than pure chance would produce?

**Formulas** (Clark & Evans, 1954,
<https://doi.org/10.2307/1931034>; edge correction
Donnelly, 1978):

$$D_{pois} = \frac{1}{2\sqrt{\lambda}}, \quad \lambda = \frac{n}{\text{area}}
\qquad D_{Donnelly} = D_{pois} + \left(0.0514 + \frac{0.0412}{\sqrt{n}}\right)\frac{\text{perimeter}}{n}$$

$$R = \frac{D_{obs}}{D_{Donnelly}}, \qquad Z = \frac{R-1}{SE(R)}$$

$R<1$ indicates clustering, $R>1$ indicates a more regular/dispersed
pattern than random, $R\approx1$ is consistent with complete spatial
randomness (CSR). The Donnelly edge correction matters because points near
the boundary of a finite sampling window systematically appear to have
farther nearest neighbours than they truly do (their real nearest
neighbour may lie just outside the window) - without it, this app's own
empirical check found a **24% false "dispersed" rate instead of the
nominal 5%** on simulated random data.

**Nearest-neighbour computation method:** the app offers two, both exact
(never approximate - verified to agree to floating-point precision) but
very different in speed. `"matrix"` builds the full pairwise distance
matrix ($O(n^2)$ time *and* memory): fine for a few thousand points, but
at $n=16{,}705$ that's a ~2.2GB matrix rebuilt on *every* Monte Carlo
simulation, taking on the order of 15-20 minutes for even 99 simulations.
`"kdtree"` (`RANN::nn2()`, $O(n\log n)$) is ~150-250x faster at a few
thousand points and the only practical choice much above that - the same
$n=16{,}705$ case with **999** simulations (10x more than the matrix
default) finishes in about 20 seconds. `"kdtree"` is the app's default;
`"matrix"` is kept available alongside it for comparison. The choice only
affects runtime, never the result.

**Worked example** (twelve fictional points forming three tight groups
scattered across a 10×10 mm field, mimicking banding / several localized
defect zones)

```{r clark-evans-example}
set.seed(3)
cx <- c(1, 8, 4); cy <- c(1, 2, 8)  # three defect-zone centers
x <- unlist(lapply(cx, function(c) rnorm(4, c, 0.25)))
y <- unlist(lapply(cy, function(c) rnorm(4, c, 0.25)))

ce <- clark_evans_test(x, y, n_sim = 499, seed = 3)
cat(sprintf("R = %.3f  (R<1 = clustered)\n", ce$R))
cat(sprintf("Asymptotic p-value: %.4f\n", ce$p_value_asymptotic))
cat(sprintf("Monte Carlo p-value: %.4f\n", ce$p_value_monte_carlo))
cat(ce$verdict, "\n")
```

```{r clark-evans-plot, fig.height=4}
print(create_spatial_scatter_plot(x, y))
```

**Interpreting the result:** $R \ll 1$ with a small p-value here confirms
the deliberately clustered toy data really is detected as clustered - both
the asymptotic and Monte Carlo p-values agree, which is what a confident
call looks like. Note that Clark-Evans only ever sees the *bounding box of
the points themselves* as the sampling window - a single tight blob just
shrinks the window to match and can look no different from CSR. Real
clustering signal comes from nearest neighbours being close *relative to
the full spread of the data*, as with the three separated groups here.

**When it can't help you:**

- Needs at least 3 points (the app enforces this), but the asymptotic test
  is known to still over-reject somewhat at small $n$ (~15% instead of 5%
  around $n=40$, checked empirically) - **trust the Monte Carlo p-value
  over the asymptotic one when they disagree**, especially for small point
  counts.
- The Monte Carlo p-value is *not* fully exact either, and the reason is
  structural: the simulated null point sets are drawn inside the observed
  data's own bounding box, but that box is tautologically as large as the
  real sample's own range, while a freshly-simulated null sample of the
  same size will (in expectation) have a *smaller* realized range - making
  the real data look artificially spread toward the edges relative to the
  simulations. Checked empirically (1000+ simulated truly-random point
  sets per $n$): at $n=15$ this gives a false "significantly dispersed"
  rate of ~18% (Monte Carlo) vs the nominal 5%, dropping to ~7-8% by
  $n=100$. Crucially, the bias is **asymmetric**: at the same $n=15$, a
  false "significantly clustered" verdict occurred in well under 1% of
  trials - i.e. *under* the nominal rate. In practice: a **clustered/
  banding** verdict from this test is trustworthy (if anything,
  conservative) even at small $n$; treat a **regular/dispersed** verdict
  below ~40 points with real skepticism, regardless of which p-value it's
  based on.
- Only tests spatial randomness in *position* - it says nothing about
  whether inclusion *size* correlates with location (combine with a
  size-colored scatter plot, available in the same tab, to check that
  visually).
- A rectangular bounding box is assumed; if your actual sampled region is
  a very different shape (e.g. an irregular polished section), the edge
  correction's accuracy degrades.

---

# Compositional Data Analysis: CLR / ILR transforms + PCA

**What it answers:** ordinary statistics (correlation, PCA, distance)
assume data lives in ordinary Euclidean space. Wt% chemistry columns are
**compositional** - constrained to sum to a constant - so they instead live
on a simplex. Applying ordinary methods directly is technically
inappropriate (the "closure problem": e.g. spurious negative correlations
are induced purely by the sum constraint, see §3). The fix is a log-ratio
transform first (Aitchison, 1986).

**CLR (centered log-ratio):**
$$clr_i = \ln(x_i) - \overline{\ln(x)} = \ln\!\left(\frac{x_i}{g(x)}\right)$$
where $g(x)$ is the row's geometric mean. Each CLR coordinate still maps
to one original element, so loadings/biplots stay directly interpretable -
but the $D$ CLR coordinates always sum to exactly 0, so their covariance
matrix is singular.

**ILR (isometric log-ratio,** Egozcue et al., 2003,
<https://doi.org/10.1023/A:1023818214614>**):** an orthonormal-basis
re-expression to $D-1$ coordinates with a non-singular covariance matrix.
Coordinate $j$ contrasts the (geometric) mean of the first $j$ parts
against part $j+1$:
$$ilr_j = \sqrt{\frac{j}{j+1}}\left(\overline{\ln(x_{1..j})} - \ln(x_{j+1})\right)$$
ILR is an *isometry* of CLR (distance- and angle-preserving), so PCA on
either basis gives identical variance-explained and scores - only the
loadings' interpretation differs (per-element for CLR, an abstract
"balance" for ILR).

**Worked example** (four fictional elements, six samples)

```{r coda-example}
df_coda <- data.frame(
  Al = c(1.2, 1.5, 0.9, 2.1, 1.8, 1.1),
  Si = c(6.5, 5.8, 7.2, 4.9, 5.5, 6.8),
  Fe = c(85, 87, 83, 88, 86, 84),
  Mn = c(0.8, 0.6, 1.1, 0.5, 0.7, 0.9)
)
parts <- c("Al", "Si", "Fe", "Mn")

clr <- clr_transform(df_coda, parts)
round(clr, 3)
cat("\nRow sums (should all be ~0):\n")
print(round(rowSums(clr), 12))

ilr_res <- ilr_transform(df_coda, parts)
round(ilr_res$ilr, 3)

pca <- compositional_pca(clr)
cat(sprintf("\nPC1 + PC2 explain %.1f%% of the (Aitchison) variance\n",
            sum(pca$var_explained[1:2])))
```

```{r coda-plot, fig.height=5}
print(create_coda_biplot(pca))
```

**Interpreting the result:** the CLR row sums confirm the transform is
correct (defining property: always ≈0). The biplot's arrows show which
elements drive the main directions of variation, in Aitchison
(log-ratio) geometry rather than raw-percentage geometry - two samples
close together in the biplot are compositionally similar even if their
raw percentages differ a lot in absolute terms.

**When it can't help you:**

- Needs **at least 3 compositional parts** and, per row, no more than a
  few genuine zeros - each zero gets replaced with a small pseudo-count
  (half the smallest positive value found), a simplified version of the
  standard approach (Martín-Fernández et al., 2003,
  <https://doi.org/10.1023/A:1023866030544>) that doesn't fully account
  for a detection-limit-per-element, since this app's data doesn't carry
  one.
- If the selected parts are only a small subset of the *true* full
  composition (e.g. 4 elements out of a 15-element analysis), the
  log-ratios are relative to that subset's own geometric mean, not the
  true whole - reasonable for comparing samples to each other, but not a
  substitute for the real full-composition analysis if it's available.

---

# Rose diagrams (directional statistics)

**What it answers:** how are inclusion *orientation/direction* values
distributed around the full circle - useful for detecting a preferred
elongation direction (e.g. from rolling), which plain histograms obscure
because they treat 359° and 1° as maximally different instead of adjacent
(Mardia & Jupp, 2000).

**Method:** direction values are binned into fixed-width sectors on a
0-360° grid (independent of the data's own range - many orientation
measurements are only recorded 0-180°, since a line has no front/back;
the empty half of the resulting rose is expected, not an error), counted
per bin, and drawn as radial bars.

**Worked example** (eight fictional directions, degrees, clustered around
20-40°)

```{r rose-example}
set.seed(4)
directions <- data.frame(dir = c(rnorm(6, mean = 30, sd = 8), 150, 165))
p_rose <- build_custom_plot(directions, "rose", x = "dir", rose_bin_width = 15)
print(p_rose)
```

**Interpreting the result:** a rose diagram is a descriptive
visualization, not a hypothesis test on its own - a visibly dominant
sector suggests a preferred orientation, but formally testing that
(e.g. a Rayleigh test of circular uniformity) isn't implemented in this
app. Use it to decide whether directional structure is worth
investigating further, not as a p-value.

**When it can't help you:** with very few points, individual
measurements dominate single sectors and the diagram can look
"directional" purely by chance - as with any histogram, don't over-read a
rose diagram built from a handful of points.

---

# References

- Aitchison, J. (1986). *The Statistical Analysis of Compositional Data*. Chapman and Hall, London.
- Anderson, T.W., Darling, D.A. (1952). Asymptotic Theory of Certain "Goodness of Fit" Criteria Based on Stochastic Processes. *Annals of Mathematical Statistics*, 23(2), 193-212. <https://doi.org/10.1214/aoms/1177729437>
- Clark, P.J., Evans, F.C. (1954). Distance to Nearest Neighbor as a Measure of Spatial Relationships in Populations. *Ecology*, 35(4), 445-453. <https://doi.org/10.2307/1931034>
- Donnelly, K. (1978). Simulations to Determine the Variance and Edge-Effect of Total Nearest-Neighbour Distance. In I. Hodder (Ed.), *Simulation Studies in Archaeology*, pp. 91-95. Cambridge University Press.
- Egozcue, J.J., Pawlowsky-Glahn, V., Mateu-Figueras, G., Barceló-Vidal, C. (2003). Isometric Logratio Transformations for Compositional Data Analysis. *Mathematical Geology*, 35(3), 279-300. <https://doi.org/10.1023/A:1023818214614>
- Jolliffe, I.T. (2002). *Principal Component Analysis*, 2nd ed. Springer. <https://doi.org/10.1007/b98835>
- Leys, C., Ley, C., Klein, O., Bernard, P., Licata, L. (2013). Detecting outliers: Do not use standard deviation around the mean, use absolute deviation around the median. *Journal of Experimental Social Psychology*, 49(4), 764-766. <https://doi.org/10.1016/j.jesp.2013.03.013>
- Liu, F.T., Ting, K.M., Zhou, Z.-H. (2008). Isolation Forest. *2008 Eighth IEEE International Conference on Data Mining*, 413-422. <https://doi.org/10.1109/ICDM.2008.17>
- Mardia, K.V., Jupp, P.E. (2000). *Directional Statistics*. Wiley. <https://doi.org/10.1002/9780470316979>
- Martín-Fernández, J.A., Barceló-Vidal, C., Pawlowsky-Glahn, V. (2003). Dealing with Zeros and Missing Values in Compositional Data Sets Using Nonparametric Imputation. *Mathematical Geology*, 35(3), 253-278. <https://doi.org/10.1023/A:1023866030544>
- Murakami, Y. (1994). Inclusion Rating by Statistics of Extreme Values and Its Application to Fatigue Strength Prediction and Quality Control of Materials. *Journal of Research of the National Institute of Standards and Technology*, 99(4), 345-351. <https://doi.org/10.6028/jres.099.032>
- Stephens, M.A. (1977). Goodness of Fit for the Extreme Value Distribution. *Biometrika*, 64(3), 583-588. <https://doi.org/10.1093/biomet/64.3.583>
- Tukey, J.W. (1977). *Exploratory Data Analysis*. Addison-Wesley, Reading, MA.
- Vode, F., Tehovnik, F., Kosec, G., Steiner Petrovič, D. (2022). Classification of Hot-Rolled Plates Using the Mahalanobis Distance of NMIs in Ti-Stabilized Austenitic Stainless-Steel Produced by Secondary Metallurgy. *Materials*, 15(2), 684. <https://doi.org/10.3390/ma15020684>
- ASTM E2283-08(2019). *Standard Practice for Extreme Value Analysis of Nonmetallic Inclusions in Steel and Other Microstructural Features*. ASTM International.
