---
title: "User Guide"
author: "Vid Kuder Marušič"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# About this guide

This document explains **how to use** the vidternary Shiny app, tab by tab:
what to click, in what order, what each tab expects as input, and what it
produces. It deliberately stays away from formulas and statistical theory -
for *why* a method works, when its result can be trusted, and a worked
numeric example of every calculation, see the companion
**Statistical Appendix** vignette (`vignette("statistical-appendix",
package = "vidternary")`). The two documents cross-reference each other
throughout.

Launch the app with:

```{r launch, eval=FALSE}
library(vidternary)
run_app()
```

# Data requirements and shared settings

## File format

Every tab that accepts data reads an **Excel file** (`.xlsx`, and on most
tabs also `.xls`) and always reads **Sheet 1 only** - there is no sheet
picker anywhere in the app, so put your data on the first sheet. Column
headers become the column names shown in every dropdown, so short,
consistent, spreadsheet-safe names (e.g. `Al2O3_Wt%`, `ECD_um`,
`StageX_mm`, `Field`, `Area_mm2`) work best.

## Uploads are per tab, not global

**There is no single "load your data once" step.** Each tab has its own
independent upload button, and a file uploaded on one tab does not appear
on any other tab. The table below shows what each tab expects:

| Tab | Upload | How many files? |
|---|---|---|
| Ternary Plots | Primary + optional Reference | exactly 1 or 2 |
| Data Comparison | one upload box | any number |
| Multiple Ternary Creator | one upload box | any number (one plot per file) |
| Hexagonal Ternary Diagram | one upload box | exactly 1 |
| Plot Builder | one upload box | any number (rows combined) |
| Extreme Value Analysis | one upload box | any number (rows combined) |
| Spatial Clustering | one upload box | any number (rows combined) |
| Compositional Analysis | one upload box | any number (rows combined) |
| Analysis Log | *(none - reads the session's activity log)* | - |

When a tab combines several files, rows get an extra `source_file` column
so you can tell datasets apart afterwards, and only the columns that are
**common to every selected file** survive the combine - if two files share
no column names, you'll see a validation error instead of a partial merge.

## Directory Settings and Cache

Every tab shares one Working Directory / Output Directory panel (native
folder pickers) at the page level, plus a Cache Management control. Any
"Save" action across the app writes into the configured Output Directory,
defaulting to an `output/` subfolder next to the app if you never pick one.
Set this once at the start of a session.

# Glossary

- **NMI** - Non-Metallic Inclusion: a small oxide/sulfide particle trapped
  in steel, the object of study throughout the app.
- **ECD** - Equivalent Circle Diameter: the diameter of a circle with the
  same area as an (irregularly shaped) inclusion, the standard way to
  express inclusion "size" as a single number.
- **Wt%** - weight percent: oxide/element composition columns that sum to
  (approximately) 100% per inclusion - this constraint is why the
  Compositional Analysis tab exists (see §8 and the Statistical Appendix).
- **Control area / field of view** - one SEM scan region; extreme value
  analysis (§6) works by taking the single largest inclusion found in each
  control area, not every inclusion in it.
- **Stage X/Y** - the microscope stage coordinates recorded for each
  inclusion, used by Spatial Clustering (§7) to test whether inclusions are
  randomly scattered or clustered (banding).
- **source_file** - a column the app adds automatically whenever a tab
  combines more than one uploaded file, recording which file each row came
  from.
- **Outlier vs. extreme value** - an *outlier* (IQR/Z-score/MAD/Mahalanobis/
  Isolation Forest, §3) is a point that looks unusual **within the dataset
  you have**; an *extreme value* (§6) is a statistical **prediction of a
  value you have not yet observed**, extrapolated from the pattern of
  per-area maxima. They answer different questions - see §9.

# Tab-by-tab instructions

## Ternary Plots

**Purpose:** the core single-plot workflow - one ternary diagram (or two,
side by side) with per-element filters and, optionally, one statistical or
multivariate outlier method.

**Steps:**

1. Upload your Primary file (and, if you want a side-by-side comparison, a
   Reference file).
2. Pick Element A, B and C from the dropdowns. Each is a multi-select - if
   you pick more than one column for a vertex, they are **summed**.
3. A filter box appears under each selected element. Type a condition like
   `> 10`, `<= 5.2`, `== 3` or `!=`; several filters combine with AND.
4. Optionally set Param 1 (point size or shape) and Param 2 (point color +
   palette). If Param 2 is a categorical column, a checklist of its
   categories appears with sample counts, letting you show/hide groups.
5. Under "🔧 Multivariate Analysis", pick **two or more numeric columns** in
   the Universal Column Selector. This selection feeds both the
   multivariate methods and the statistical filters below it.
6. Tick **at most one** of Mahalanobis / Isolation Forest / IQR / Z-score /
   MAD - checking one automatically unchecks any other, since the app only
   ever applies one filter method per plot.
7. Choose an output format and point size, then click **Save Plot 1 / Plot
   2 / Both**.

The two live previews update automatically as you change elements or
filters - no button needed for previewing, only for saving.

**Tips & gotchas:**

- IQR/Z-score/MAD are *not* per-element: they run on the Universal Column
  Selector set and flag a row if **any** of those columns is an outlier
  (positive/high-value side only).
- "📋 Copy Settings from Dataset 1" copies every element/filter/palette/
  filter-method choice to Dataset 2 in one click - handy once you've tuned
  Dataset 1 and want a matching Dataset 2 plot.
- Changing which columns are selected for an element rebuilds that
  element's filter box, so any text you'd typed into it is lost.
- Saved images go straight into the Output Directory root (not a
  timestamped subfolder), named
  `charge_<originalfilename>_<timestamp>.<format>`.

## Multiple Ternary Creator

**Purpose:** batch-generate one ternary plot per uploaded file, all using
one shared set of Element A/B/C and optional-parameter settings. There is
no outlier filtering here - use the Ternary Plots tab per file if you need
that.

**Steps:**

1. Upload as many files as you like. Column dropdowns populate from the
   **first** file only, so all files should share the same column names.
2. Pick an output folder name and format.
3. Choose Element A/B/C (+ optional filters/Param 1/Param 2), exactly as on
   the Ternary Plots tab.
4. Click **Create & Save** - every file is plotted and saved immediately,
   no preview step.

Errors on individual files are collected and reported together at the end;
one bad file doesn't stop the rest of the batch. Output goes to
`<OutputDir>/<folder_name>_<timestamp>/`.

## Hexagonal Ternary Diagram

**Purpose:** composites six triangular ternary plots that share one
central axis into a single hexagonal image, from one uploaded file.

**Steps:**

1. Upload one file.
2. Fill in **all seven** element-slot dropdowns (labeled A-G to match the
   legend diagram shown on the tab) - every slot is required, and slot 3
   (C) is the shared central axis for all six sub-triangles.
3. Click **Generate** ("Ustvari heksagonalni diagram") to preview.
4. Click **Save** ("Shrani v izhodno mapo") to write the final image to
   `<OutputDir>/<folder_name>_<timestamp>/`.

If any of the seven slots is left empty, both buttons simply show an error
and do nothing - there's no way to render a partial hexagon.

## Plot Builder

**Purpose:** a generic chart builder - violin, box, bar (counts),
histogram, scatter, or rose diagram - from any column in your data, with a
save/load preset system for repeat use.

**Steps:**

1. Upload one or more files; pick which of them to include via the dataset
   selector (defaults to all).
2. Choose a chart type. The axis controls rebuild for each type:
   - **Violin / Box:** X = group column (any type); Y = one or more
     numeric columns (selecting 2+ compares several elements side by side
     within each X group, and overrides Color-by).
   - **Bar (counts):** X = category column, plus a "categories to show"
     selector (drawn from Color-by if set, else from X), plus a
     show-percentages toggle.
   - **Histogram:** one numeric variable + a bin-count field (default 30).
   - **Scatter:** X and Y, both numeric.
   - **Rose diagram:** a direction column in degrees + a bin width in
     degrees (default 10).
3. Optionally set Color-by (categorical columns only) and Log-X/Log-Y.
4. The plot re-renders automatically as you change any control - there is
   no "generate" button.
5. Use **Download** for a PNG, or **Save/Load/Delete Preset** to store this
   exact configuration under a name for later sessions (presets are saved
   to disk, so they persist across app restarts).

**Tips & gotchas:**

- For Violin/Box, the Y selector starts with **nothing chosen** - the plot
  stays blank until you pick at least one Y column (or load a preset that
  sets one). This is expected, not a bug.
- Loading a preset can take a brief moment to fully apply, since the chart
  type must rebuild its axis controls before the axis values themselves
  can be restored.
- Combining multiple files keeps only columns common to all of them.

## Extreme Value Analysis

**Purpose:** fits a Gumbel extreme-value model to per-control-area maximum
inclusion sizes (Murakami's method) and predicts the largest inclusion
expected in a larger inspection area. See the Statistical Appendix for the
formula and the meaning of "return period."

**Steps:**

1. Upload one or more files (rows combined).
2. Pick the Area column (auto-detected if a column name starts with
   "area").
3. Choose how control areas are grouped:
   - Leave **"use manual groups"** unchecked and pick a real Field/Frame ID
     column (auto-detected if named `field`) - each distinct value is one
     physical control area.
   - Or check it and pick a number of groups (default 20) to split the
     data into arbitrary equal-size groups when you have no real field ID.
4. Click **Fit Extreme Value Model**.
5. Once fitted, set the **Return Period (T)** - the prediction updates live
   as you change it, no re-fit needed.
6. Download the plot (PNG) or the raw block-maxima table (XLSX).

**Tips & gotchas:**

- Needs **at least 3 groups** with valid data, or the fit is blocked.
- With a real Field/Frame ID, T has a physical meaning (T=100 = an area
  100x one field of view); with the manual-groups fallback, T is a purely
  statistical multiple with **no physical area interpretation** - the app
  flags this explicitly.
- If the Anderson-Darling goodness-of-fit test rejects the single-Gumbel
  assumption, a warning box appears; the fit line is still drawn, but the
  tail prediction should be treated cautiously (it may indicate a mixed
  inclusion population or one outlier control area skewing the fit).

## Spatial Clustering

**Purpose:** the Clark-Evans nearest-neighbour test, checking whether
inclusion positions are randomly scattered, clustered (e.g. banding), or
more regular than random.

**Steps:**

1. Upload one or more files (rows combined).
2. Pick the X and Y stage-coordinate columns (auto-detected if named like
   `stage...x` / `stage...y`).
3. Optionally pick a column to color the scatter plot by (this does not
   affect the statistical test itself).
4. Leave **Nearest-neighbour method** on "k-d tree" (the default) unless
   you have a specific reason to compare against "Distance matrix" - both
   give identical results, but the distance-matrix method can take many
   minutes on datasets above a few thousand points, while k-d tree stays
   fast into the tens of thousands.
5. Adjust **Monte Carlo simulations** if you want a finer-grained p-value
   than the default 999 (k-d tree) / 299 (distance matrix) - more
   simulations means a smaller minimum reportable p-value, at the cost of
   runtime (only meaningfully slow with the distance-matrix method).
6. Click **Analyze Spatial Pattern**.

**Outputs:** R statistic, both p-values (asymptotic and Monte Carlo),
verdict text, a colored scatter plot, a nearest-neighbour-distance
histogram, a summary table (including which method and simulation count
were used), and three downloads (scatter PNG, histogram PNG, per-point NND XLSX).

**Tips:** when the two p-values disagree, trust the Monte Carlo one,
especially with fewer points. Below ~40 points, treat a **clustered**
verdict as trustworthy but a **dispersed/regular** verdict with real
skepticism - see the Statistical Appendix for why the two directions are
not equally reliable at small sample sizes.

## Compositional Analysis

**Purpose:** CLR/ILR log-ratio transforms and PCA for Wt% chemistry
columns, which sum to a constant and therefore need a log-ratio transform
before ordinary PCA is statistically valid.

**Steps:**

1. Upload one or more files (rows combined).
2. Pick the compositional columns in "Parts to include" (auto-preselected
   if three or more columns contain "wt%" in their name).
3. Click **Transform & Run PCA**.

**Outputs:** a CLR biplot, a variance-explained table, and two loadings
tables side by side - CLR basis (one row per element, directly
interpretable) and ILR basis (one row per abstract "balance," not per
element) - plus downloads for the CLR-transformed data, ILR-transformed
data, and the biplot image.

**Tips & gotchas:**

- Requires **at least 3 selected parts** and enough complete rows (rows
  with any missing value among the selected parts are dropped first).
- Zeros are silently replaced with a small pseudo-count before
  log-transforming (log(0) is undefined) - there is no toggle to disable
  this.
- The ILR loadings depend on the **order** you selected the parts in -
  reordering them changes what each `ilr_j` balance represents, even
  though the CLR results and PCA variance-explained stay identical either
  way (this is expected - see the Statistical Appendix's note on ILR being
  an isometry of CLR).

## Analysis Log

**Purpose:** a searchable, filterable record of what happened during the
current session (uploads, saves, analyses, errors) across every tab.

**Steps:** the log fills itself automatically - nothing to configure.
Use the Level dropdown (All/INFO/WARNING/ERROR) and the search box to
narrow it down, and **Clear / Save to File / Export** to manage it.

**Tips & gotchas:**

- The log is in-memory only - restarting the app clears it. "Save to File"
  (writes into the Output Directory) is the only way to keep a permanent
  copy.
- "Save to File" and "Export" both save exactly what's currently visible -
  if you've narrowed the view with a Level filter or search, you'll save a
  partial log, not the full session history.
- The Log Statistics counts (top of the tab) always reflect the *entire*
  session log, even while a filter narrows what's shown below - so the
  visible entry count and the stats box can legitimately disagree.

# Which method should I use?

| I want to... | Go to |
|---|---|
| Plot three elements against each other as a classic ternary diagram | **Ternary Plots** |
| Produce the same ternary plot, unchanged, across many files | **Multiple Ternary Creator** |
| Compare descriptive stats/correlations across several datasets side by side | **Data Comparison** |
| See six related ternary views sharing one axis at a glance | **Hexagonal Ternary Diagram** |
| Build a violin/box/bar/histogram/scatter/rose chart from any column | **Plot Builder** |
| Flag unusual values in one or two elements | **Ternary Plots** -> IQR / Z-score / MAD |
| Flag unusual points across several elements at once | **Ternary Plots** or **Data Comparison** -> Mahalanobis or Isolation Forest |
| Predict the largest inclusion in a bigger inspection area than you sampled | **Extreme Value Analysis** |
| Check whether inclusions are clustered/banded in space | **Spatial Clustering** |
| Run PCA on Wt% chemistry without breaking the sum-to-100% constraint | **Compositional Analysis** |
| Review or save what happened this session, or debug an error | **Analysis Log** |

For the choice *between* the five outlier-flagging methods (IQR, Z-score,
MAD, Mahalanobis, Isolation Forest) and how to judge whether a flagged
point is real, see the Statistical Appendix's per-method "when it can't
help you" notes.

# A typical end-to-end workflow

Suppose you've collected SEM/EDS data for five steel heats, one Excel file
per heat, each with columns for oxide composition (`Al2O3_Wt%`,
`SiO2_Wt%`, `CaO_Wt%`, `MgO_Wt%`, `MnO_Wt%`), `ECD_um`, stage position
(`StageX_mm`, `StageY_mm`), a `Field` ID, and `Area_mm2`. A reasonable path
through the app:

1. **Ternary Plots** - upload heat 1 as Primary, plot
   `Al2O3-SiO2-CaO`, apply a MAD filter over the five oxide columns to spot
   obviously anomalous inclusions, save the figure.
2. **Multiple Ternary Creator** - once the settings from step 1 look
   right, upload all five heat files and generate the same ternary plot
   for each in one batch.
3. **Data Comparison** - upload all five files, compute descriptive
   stats and a correlation heatmap, then run Mahalanobis distance with
   "Self" as reference on each heat to see which one has the most unusual
   inclusion population.
4. **Plot Builder** - build a box plot of `ECD_um` grouped by `Field` (or
   `source_file`, i.e. heat) to compare inclusion size distributions
   visually; save this configuration as a preset for next time.
5. **Spatial Clustering** - using `StageX_mm`/`StageY_mm`, test whether
   inclusions are randomly scattered or clustered (banding) within each
   heat.
6. **Extreme Value Analysis** - using `Area_mm2` grouped by `Field`, fit a
   Gumbel model and predict the largest inclusion expected in, say, an
   area 100x a single field of view - relevant for fatigue-critical
   applications where the largest defect matters most.
7. **Compositional Analysis** - run CLR/ILR + PCA on the five oxide
   columns to see which oxides drive the main compositional variation,
   properly accounting for the sum-to-100% constraint.
8. **Analysis Log** - review the session's log and save it alongside your
   results for reproducibility.

Not every analysis is needed for every dataset - pick the tabs relevant to
the question you're actually asking (see §5 above).

# Troubleshooting / FAQ

**My Plot Builder chart is blank after I pick Violin or Box.**
Pick at least one column in the Y selector - it starts empty by design
(see §4.4).

**The Bar (counts) "categories to show" selector isn't there.**
It only appears once a chart type/column selection makes it relevant;
check that you've picked an X column (and, if you want it scoped to
groups, a Color-by column).

**Clicking Generate on the Hexagonal Ternary Diagram tab does nothing.**
All seven element slots (A-G) must have at least one column selected -
there's no partial-hexagon mode.

**Extreme Value Analysis or Compositional Analysis refuses to fit/run.**
EVA needs at least 3 control-area groups with valid data; Compositional
Analysis needs at least 3 selected parts and enough complete rows (rows
with no missing values among the selected columns) - both show a
validation message explaining which condition failed.

**The correlation heatmap disappeared on Data Comparison.** It only draws
for 1 or exactly 2 selected datasets; with 3+ datasets selected, the
side-by-side comparison table still works, but narrow your selection to 2
if you need the heatmap.

**I changed the Mahalanobis lambda/omega settings and they changed on
another tab too.** The Ternary Plots and Data Comparison tabs share the
same lambda/omega/threshold-mode inputs by design - they are one setting,
not two independent ones.

**A filter box I typed into disappeared.** Changing which column(s) are
selected for an element (Ternary Plots) rebuilds that element's filter
box, clearing any previously typed text.

**Combining multiple files dropped some of my columns.** Tabs that combine
files (Plot Builder, Extreme Value Analysis, Spatial Clustering,
Compositional Analysis, Data Comparison) keep only columns common to every
selected file. If two files have no columns in common at all, you'll see a
validation error instead of a partial merge.

# Where to go next

For formulas, assumptions, statistical significance thresholds, and a
live worked numeric example of every method mentioned above, see
`vignette("statistical-appendix", package = "vidternary")`.
