---
title: "Introduction to Vidternary"
author: "Vid Kuder Marušič"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Vidternary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction to Vidternary

Vidternary is a comprehensive R package for creating ternary plots with advanced filtering and multivariate analysis capabilities. This vignette provides a quick introduction to the package's main features.

## Installation

```{r install, eval=FALSE}
# Install required packages first
install.packages(c("shiny", "openxlsx", "ggplot2", "DT", "corrplot", "GGally", "Ternary", "PlotTools", "robustbase", "isotree"))

# Install from GitHub
devtools::install_github("vidkudermarusic/vidternary")

# Or load locally for development
source("R/app.R")
```

## Basic Usage

### Interactive Application
```{r basic-usage, eval=FALSE}
library(vidternary)

# Run the full interactive Shiny application
run_app()

# Or create app object for custom deployment
app <- create_app()
shiny::runApp(app)
```

### Programmatic Usage
```{r programmatic-usage, eval=FALSE}
library(vidternary)

# Initialize packages (loads all dependencies)
initialize_packages()

# Create a ternary plot programmatically
result <- general_ternary_plot(
  xlsx_file = "data.xlsx",
  element_A = list(col = "SiO2", filter = NULL),
  element_B = list(col = "Al2O3", filter = NULL),
  element_C = list(col = "Fe2O3", filter = NULL),
  output_dir = "output/",
  color_palette = "viridis"
)
```

## Key Features

- **Ternary Plot Generation**: Create publication-ready ternary plots
- **Advanced Filtering**: Statistical (IQR, Z-score, MAD - positive/high-value outliers only) and multivariate filtering methods (Mahalanobis, Isolation Forest)
- **Hexagonal Ternary Diagrams**: Composite 6 triangular plots sharing a central element/combination into one image
- **Plot Builder**: Build custom violin/box/bar/histogram/scatter charts from any column, with saved presets
- **Interactive Interface**: User-friendly Shiny application
- **Export Capabilities**: Multiple output formats
- **Modular Architecture**: Easy to extend and maintain

## Getting Help

For more information, see the package documentation or visit the GitHub repository.
